Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
edits up to section 4, will start rest soon

\subsection{Avoiding polymorphic lift}

@@

 val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
 val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
 val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
 val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)

end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}

@@

  let buildsimp cast e f1 f2 = fun e1 e2 -> 
     match (e1,e2) with
        | ({st = Some e'}, _) when e = e' -> e2
        | (_, {st = Some e'}) when e = e' -> e1
        | ({st=Some n1}, {st=Some n2})-> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))

  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
  let leq e1 e2 = 
     match (e1,e2) with
         ({st=Some n1},{st=Some n2})-> bool (R.leq n1 n2)
        | _ -> pdyn (C.leq (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (abstr ef) (abstr ea))

@@

The type |repr| here literally expresses the type equation at the
beginning of this section. The function |abstr|, as in the previous
subsection, extracts the future-stage computation from the result of
partial evaluation.  The function |pdyn| does the converse. As in
section~\ref{S:PE-problem}, we build future-stage computations at the
same time as we
build present-stage ones, if any. Therefore, we avoid the need for a
polymorphic `lift' function. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage 
respectively of the implicit monoid and ring structure of the 
underlying domain.
The function |buildsimp| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function also checks if one argument is static
and is the neutral element of the monoid structure 
\jacques{Comment on the various nice generic things in the implementation,
which are a little silly in this small version, but would really pay off
in something bigger.  See the abstract interpretation hard at work!
See the monoid-driven simplifications!}
\oleg{I'm inclined to ditch buildsimp. We miss an opportunity here:
  when multiplying anything by 0, the result is zero!}

When doing partial evaluation on |fix|, we made a deliberate decision to `go all
the way' -- so long as the computation can proceed at the PE stage,
we do that rather than residualizing. In the
accompanying source code, we have also implemented an alternative
where |fix| is unrolled only once and then we residualize. We prefer the
first approach, even though it may cause
non-termination at the PE stage -- but \emph{only} if the computation
we are partially evaluating is non-terminating, and enough static
information is available to trigger this non-termination.  Our
treatment of |fix| may lead to code bloat in the residual
program. That is an issue investigated elsewhere (cite our PEPM2006
paper).

Given the above Symantics, our running test now evaluates
\begin{code}
  let module E = EX(P) in E.test1 ();;
\end{code}
to
\begin{code}
- : ('a, bool, bool) P.repr = 
   {P.st = Some true; P.dy = .<(true)>.}
\end{code}
It is instructive to compare this result with that of the |C|
evaluator. We observe the beta-redex that was apparent there is now
performed. The result is the static value |true|.
More interesting is the partial evaluation of testpowfix7. The |C|
interpreter evaluates |testpowfix7| to the future-stage code with many
apparent beta-redexes.
\begin{code}
  let module E = EX(C) in E.testpowfix7;;
 - : ('_a, int -> int) C.repr =
.<fun x_1 ->
   (((fun x_2 ->
       let rec self_3 =
        fun n_4 ->
         ((fun x_5 -> if (x_5 <= 0) then 1 
                      else (x_2 * (self_3 (x_5 + (-1)))))
           n_4) in
       self_3) x_1) 7)>.
\end{code}
The result of the |P| interpreter is different:

\begin{code}
  let module E = EX(P) in E.testpowfix7;;
 - : ('_a, ('_a, int, int) P.repr -> ('_a, int, int) P.repr, 
      int -> int) P.repr = 
 {P.st = Some <fun>;
  P.dy = .<fun x_1 -> 
     (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1))))))>.}
\end{code}

which is what one would have expected from the power example.

Unlike the GADT approach in the previous subsection, there is no
longer any seemingly partial pattern-matching. The pattern-match only
occurs in PE (to check if the code is static or dynamic), it is only a
phase rather than a type tag match, and the pattern match is
\emph{patently} exhaustive. That is, totality is assured and it is
assured \emph{syntactically}, in the way that is apparent to the
compiler.

One should note that the partial evaluator has fully reused the code
of the compiler and the interpreter. It is truly the `composition' of the
interpreter and the compiler.  \jacques{Perhaps cite a paper of Thiemann
here, cogen in 6 lines?}

\jacques{Comment on online/offline here}
The PE in incope is really a mixture of online and off-line; it looks 
online, but it has a sort of "just in time" binding-time analysis built in.

@@

\oleg{where to put the following? Should we even mention this?
If not, we should still keep this in comments.}
If we just add a method 'lift' to Symantics with a type
|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
MetaOCaml: CSP)
then |Symantics repr| implies |Functor repr|. We can write then
\begin{code}
  instance Semantics repr => Functor repr where
  -- fmap :: (a->b) -> repr a -> repr b
  fmap f ra = app (lift f) ra
\end{code}

@@

>>>>

\subsection{Avoiding polymorphic lift}
\label{S:PE-lift}

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}

@@

  let build cast f1 f2 = fun e1 -> fun e2 -> match (e1,e2) with
                   ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
                 | _ -> pdyn (f2 (abstr e1) (abstr e2))
  let buildsimp_monoid cast e f1 f2 = fun e1 e2 -> match (e1,e2) with
                 | ({st = Some e'}, _) when e = e' -> e2
                 | (_, {st = Some e'}) when e = e' -> e1
                 | _ -> build cast f1 f2 e1 e2
  let buildsimp_ring cast zero one f1 f2 = fun e1 e2 -> match (e1,e2) with
                 | ({st = Some e'}, _) when e' = zero -> e1
                 | (_, {st = Some e'}) when e' = zero -> e2
                 | _ -> buildsimp_monoid cast one f1 f2 e1 e2

  let add e1 e2 = buildsimp_monoid int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp_ring int 0 1 R.mul C.mul e1 e2
  let leq e1 e2 = build bool R.leq C.leq e1 e2
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (abstr ef) (abstr ea))

@@

The type |repr| here literally expresses the type equation at the
beginning of this section. The function |abstr|, as in the previous
subsection, extracts the future-stage computation from the result of
partial evaluation.  The function |pdyn| does the converse. As in
section~\ref{S:PE-problem}, we build future-stage computations at the
same time as we
build present-stage ones, if any. Therefore, we avoid the need for a
polymorphic `lift' function. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage 
respectively of the implicit monoid and ring structure of the 
underlying domain.
The function |build| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function |buildsim_monoid| also checks if one argument
is static and is the neutral element of the monoid structure.
|buildsimp_ring| also worries about the full ring structure rather than
just the monoid structure.  These are put here to illustrate how we 
would go about writing general algebraic simplification routines that
would mimic various compiler optimizations, and these would be quite
beneficial in a large language with more base types.

When doing partial evaluation on |fix|, we made a deliberate decision to `go
all the way' -- so long as the computation can proceed at the PE stage,
we do that rather than residualizing. In the
accompanying source code, we have also implemented an alternative
where |fix| is unrolled only once and then we residualize. We prefer the
first approach, even though it may cause
non-termination at the PE stage -- but \emph{only} if the computation
we are partially evaluating is non-terminating, and enough static
information is available to trigger this non-termination.  Our
treatment of |fix| may lead to code bloat in the residual
program. The techniques of~\citet{SwadiTahaKiselyovPasalic2006} can
be used to deal with this issue.

Given the above Symantics, our running test now evaluates
\begin{code}
let module E = EX(P) in E.test1 ();;
\end{code}
to
\begin{code}
- : ('a, bool, bool) P.repr = 
   {P.st = Some true; P.dy = .<(true)>.}
\end{code}
It is instructive to compare this result with that of the |C|
evaluator. We observe that the beta-redex has now been 
performed. The result is the static value |true|.
More interesting is the partial evaluation of testpowfix7. The |C|
interpreter evaluates |testpowfix7| to the future-stage code with many
apparent beta-redexes.
\begin{code}
let module E = EX(C) in E.testpowfix7;;
 - : ('_a, int -> int) C.repr =
.<fun x_1 -> (((fun x_2 ->
  let rec self_3 =
   fun n_4 ->
    ((fun x_5 -> if (x_5 <= 0) then 1 
                 else (x_2 * (self_3 (x_5 + (-1)))))
      n_4) in
  self_3) x_1) 7)>.
\end{code}
The result of the |P| interpreter is different:

\begin{code}
let module E = EX(P) in E.testpowfix7;;
 - : ('_a, ('_a, int, int) P.repr -> ('_a, int, int) P.repr, 
      int -> int) P.repr = 
 {P.st = Some <fun>;
  P.dy = .<fun x_1 -> 
     (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1))))))>.}
\end{code}
as expected from the power example.

Unlike the GADT approach in the previous subsection, there is no
longer any seemingly partial pattern-matching. All pattern-matching
occurs in the partial evaluator only (to check if the current value 
is static or dynamic), and is 
phase rather than a type tag match. Furthermore the matching
\emph{patently} exhaustive. That is, totality is assured and it is
assured \emph{syntactically}, in a way that is apparent to the
compiler.

One should note that the partial evaluator has fully reused the code
of the compiler and the interpreter. It is truly the `composition' of the
interpreter and the compiler.   This is a simpler situation than
in \citet{SperberThiemann:TwoForOne} where a partial evaluator and
a compiler are `composed', but the general ideas are similar.

Lastly, we should comment on the approach taken in the partial evaluator.
It is an online, typed, polyvariant partial evaluator.  It has no
termination analysis, and as written may diverge (but only on divergent
programs) as we have decided to aggressively unroll loops.

@@

@@

\section{Lifting and Functor}

It is perhaps worthwhile noting that 
if we were to add a method 'lift' to |Symantics| with type
|lift :: a -> repr a| (the polymorphic lift of \citet{xi-guarded}, 
of MetaML and
MetaOCaml: CSP)
then |Symantics repr| implies |Functor repr|.  This can be seen from
\begin{code}
instance Semantics repr => Functor repr where
  -- fmap :: (a->b) -> repr a -> repr b
  fmap f ra = app (lift f) ra
\end{code}
Of course, as we have mentioned earlier in~\S\ref{S:PE-lift}, 
addin such a polymorphic lift would appear to introduce more
problems than it solves.
